---
title: "Tristan Swartout Lab 5 assignment - Point pattern analysis"
output:
  html_document:
    df_print: paged
---

```{r, warning=F, error=F, message=F}
rm(list=ls())

require(spatstat)
require(tidyverse)
require(terra)

```


# Challenge 1 (3 points)

Create a simulation window on a landscape stretching from 0 to 100 units in both the x and y directions. Now simulate the distribution of an organism on this landscape using a Poisson process with a lambda value of 0.01. Plot this organism's distribution. How many organisms are there on your landscape? What is the expected number of organisms that you should find on this landscape based on your simulation parameters? Why?

First let's make an extent map


```{r}
set.seed(23)

Simul = rpoispp(lambda = 0.01, nsim=1, win=owin(c(0,100),c(0,100)))
plot(Simul)
npoints(Simul)
```


Answer 1:

After the simulation we have 101 randomly placed individuals. Based off expectation of what lambda in a rpoispp is defined as (the number of points per unit area) we should expect 100 individuals because a 0.01 would indicate a 0.01 portion of the total area which in this case is 10,000 units squared which would give you a value of 100. Why there is 101 points and not exactly 100 is likely due to it being a random generator. 


# Challenge 2 (3 points)

Verify that there no linear or quadratic pattern in the distribution of your organisms using fitted Poisson process models. Show the outcome of this analysis and explain your conclusion.


First lets check for CSR using chi squared
```{r}
q = quadratcount(Simul, nx=4, ny=4)
plot(Simul)
plot(q, add=T)

quadrat.test(Simul, nx=4, ny=4, method='Chisq')
```

Indication it is random


Now let's look at the Ripleys L test to see how it appears

```{r}
Lmodtrans = Lest(Simul, correction='translate')
plot(Lmodtrans, .-r~r)
```
Now with CI's

```{r}
Lmod2 = envelope(Simul, Lest, nsim=99, rank=1, correction='translate', global=F)
plot(Lmod2, .-r~r, shade=c('hi', 'lo'), legend=F)
```
Once again indication it is random

Finally lets look for inhomogeneous point process by running models where we account for a homogeneous point process (i.e., the null), then a linear model, and finally a quadratic model.


```{r}
Null = ppm(Simul ~1)
Mxy = ppm(Simul ~ x + y)
Mxy2 = ppm(Simul ~polynom (x,y,2))
```

Then lets put out an AIC table


```{r}
data.frame(model = c('int', 'xy', 'xy2'),
           AIC = c(AIC(Null), AIC(Mxy), AIC(Mxy2)))
```


Answer 2:

Not surprising our random simulation model passed the tests of randomness for a chi-squared test and Ripleys L test indicating no spatial aggregation. However, when running linear and quadratic effect models we found that the AIC value for the linear model was very similar and competing to the null model (i.e., randomness). This indicates that the random point simulation may follow some coordinate linear pattern on the extent grid. 


# Challenge 3 (14 points)

I have simulated the distributions of 2 critters and provided their x-y coordinates. One of them is a non-territorial predator. The second is a critter that is preyed upon by the predator and exhibits a fear response to the predator's presence. Use the tools you learned in the lab to explore the distributions of these two critters. Identify which one is the predator and which is the prey, and provide code and a thorough justification supporting your conclusion.


```{r}
#Use this code chunk to get you started.

predPreyWin = owin(xrange = c(0, 100), yrange = c(0, 100))

critter1 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species1.csv') %>% 
  mutate(org = 'critter1')

critter2 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species2.csv') %>% 
  mutate(org = 'critter2')
```

Lets plot critter 1

```{r}
pppcrit1=ppp(x=critter1$x, y=critter1$y, window = predPreyWin)
plot(pppcrit1)
```

Let's examine if critter 1 is aggregated and not following CSR

```{r}
Crit1plot = envelope(pppcrit1, Lest, nsim=99, rank=1, correction='translate', global=F)
plot(Crit1plot, .-r~r, shade=c('hi', 'lo'), legend=F)
```


Now lets plot critter 2

```{r}
pppcrit2=ppp(x=critter2$x, y=critter2$y, window = predPreyWin)
plot(pppcrit2)
```


Now let's see if there is aggregation and not following CSR


```{r}
Crit2plot = envelope(pppcrit2, Lest, nsim=99, rank=1, correction='translate', global=F)
plot(Crit2plot, .-r~r, shade=c('hi', 'lo'), legend=F)
```

Based off this both species appear to follow CSR and not clustering 


Now lets overlay the points

```{r}
Comb = ppp(c(pppcrit1$x,pppcrit2$x),c(pppcrit1$y,pppcrit2$y),marks = factor(c(rep("A",pppcrit1$n),rep("B",pppcrit2$n))), window = predPreyWin)
plot(Comb)
```

Lets first run a ripley L test without accounting for bivariate points; ignoring the distribution of critter b


```{r}
Model1 = envelope(Comb, Lest, nsim=99, rank=1, i='A', global=F)

plot(Model1, .-r~r, legend=F)
```


And now let's run a multi model utilizing i to account for the distribution of both critters we will make i critter A

```{r}
Model1iA = envelope(Comb, Lcross, nsim=99, rank=1, global=F, i='A', simulate = expression(rlabel(Comb)))
plot(Model1ij, .-r~r, legend=F)

```

Based off this, this indicates there is a form of clustering occurring where critter A is in relation to critter B


Now lets do the opposite of what we did before and make i critter B

```{r}
Model1iB = envelope(Comb, Lcross, nsim=99, rank=1, global=F, i='B', simulate = expression(rlabel(Comb)))
plot(Model1iB, .-r~r, legend=F)

```

Now it's a whole different story, critter B is not clustered due to critter A and follows CSR

Answer 3:

Based off these findings Critter A (critter 1) is clustering and not following a random distribution due to critter B (critter 2). This would indicate to me that critter B (critter 2) is the predator and critter A (critter 1) is clustering together to avoid predation events. I made this assumption based off the typical behavior of many species of prey that will cluster in numbers to avoid predation because more individuals together will result in a lower probability of any given individual being predated. 